---
title: "Regression Models course project, Data science specialization"
author: "Ilyas Aaqaoui"

output: pdf_document
---

#Executive Summary   
Motor trend is a magazine about the automobile industry. They are interested in studying the relationship between a set of variables and fuel consumption, measured by Miles per Galon. After conducting a study: cleaning, exploring and applying regression models on the data set provided, we can say that manual transmission is better than automatic transmission in terms of fuel consumption. In the following we will showcase this result and provide a measurment of how much manual is better than automatic.   

#Study

The data set in hand is called "mtcars", extracted from the 1974 Motor Magazine. it has in total 32 cars (observations) and 11 variables. we interested in the mileage per galon (MPG) as the outcome dependant on the other variables (regressors), in particular, the transmission mode:
  – is a manual or automatic transmission better for MPG?
  – quantify the MPG difference between auto and manual.
     
  
The data has, besides MPG, 10 other variables, classical measures in the auto industry: 
 number of cylinders, displacement, horsepower, weight, transmission etc.
 
Let's start with some cleaning.  

##Cleaning Data
We load the data and convert some variables to factors, as they should be!

```{r}
data(mtcars)
mtcarsOld = mtcars
mtcars$am   = factor(mtcars$am)
mtcars$gear = factor(mtcars$gear)
mtcars$cyl  = factor(mtcars$cyl)
mtcars$carb = factor(mtcars$carb)
mtcars$vs   = factor(mtcars$vs)
```

For more easyness, let's change the am levels names to more significative ones:
```{r}
levels(mtcars$am) = c("Automatic", "manual")
```

###Packages Loading
```{r results='hide'}
library(ggplot2)
library(caret)
library(dplyr)
library(bitops)
library(RCurl)
```

   


##Exploratory Analysis
The objective of this part is to explore data and relationship between variables for better modelling.

###Relationship/Correlation between variables
 We plot mpg against each of the other variables(see appendix - fig.1) to understand how they are correlated with it (just as we did with the Swiss feritility data). We find out that most of these variableas are highly correlated with mpg, individually (mpg with each variable separately).   

Let's now try to find the correlation between regressors, excluding the outcome, in order to find which variable should not be included in the model as predictor. 

```{r}
c=cor(mtcarsOld)
corIndices = findCorrelation(c, cutoff = 0.75) ; print(corIndices)
```
We find that variables of indices 2, 3, 1, 9 which are  cyl, disp, mpg, and am are highly correlated. mpg is the output variable. am is the variable under study so we will include both of mpg and am while cyl, and disp should preferably not be included in the model together (just like the Education and Examination variables in Swiss data). We will leave the step function to fine tune the selection process.  

### The effect of transmission type
Also, to compare the effect of the type of transmission of the fuel consumption, we plot the different distributions of mpg vs transmission type(see appendix - fig.2). The figure shows that manual transmission cars tend to consume less fuel.  

##Regression Analysis 

 We build a base regression model that include all variablea as predictors for the MPG then we use the step function to build further linear models to select the most approprate variables as predictors while eliminating those that are not very significant to the model.

```{r, results='hide'}
baseModel = lm(mpg ~., data= mtcars)
bestModel = step(baseModel, direction = "both")
```
As we can see, the algorithm has selected cyl, hp, wt and am as predictors.      
```{r}
summary(bestModel)
```   

but let's check the strength of our model through residuals:
###Residuals and diagonistics  

To check residuals for normaility and homoscedasticity, we plot the residuals(see appendix - fig.3). We find that the residuals are normally distributed and homoscedastic.   

##Results of the Model Chosen

The model we used for the regression included the cylinders, horsepower weight and transmission.

###Variability of the data measurment
This model, bestModel, has successflly explained more than 83% of the variablilty of the data.

###Model coefficients   
**Transmission control**   
Miles per gallon increases by a factor of 1.8 (1.8) with manual transmission.   

**mpg vs hp**   
Miles per gallon decreases by a factor of 0.03 (-0.03) as horspower increases.   

**mpg vs hp**   
Miles per gallon decreases by a factor of 2.5 (-2.5) for every increase of 1000lb in the weight.   

**mpg vs cyl6/cyl8**   
Miles per gallon decreases by a factor of 3.03 (-3.03) for 6 cylinders and by a factor of 2.16 (-2.16) for 8 cylinders. Which is normal since more cylinders mean more fuel injection to the pistons chambers. 

###Intercept   
The intercept is at 33.7 mpg.   

### P-value   
The overall p-value is very small (1.506 x 10 ^ -10), we then reject the null hypothesis that variables are not significant.  

##Conclusion   
####"Which one is better for MPG”   
As illustrated in this study, the manual transmission is better for fuel consumption, miles per gallon. This is demonstrated by the best fit model which estimates the decrease of fuel consumption by a factor of 1.8 for manual transmission.   

####"Quantifying the difference between automatic and manual transmissions"    
Fuel consumption, mpg, for manual transmission increases by a factor of 1.8 over automatic transmission. 
```{r}
manualData = mtcarsOld[ mtcarsOld$am == 0, 1]
automaticData = mtcarsOld[mtcarsOld$am == 1, 1]
manualCI = t.test(manualData, mu = 0)$conf.int
automaticCI = t.test(automaticData, mu = 0)$conf.int
print(manualCI);print(automaticCI)
```
The 95% confidence interval for mpg for automatic transmission (15.3, 19)  while it is (20.67, 28.12) for the manual transmission. This is an evidence that manual transmission consumes less fuel. Manual transmission travels around 7.24 more miles per gallon, (24.39mpg), than automatic transmission (17.15 mpg). 

##Appendix
###Exploratory features plot   
```{r appendix-fig1, fig.width=7, fig.cap="Figure 1. Featues plot for the dataset",  echo= FALSE}
#featurePlot(x=mtcars[,c("cyl", "disp")], y = mtcars$mpg, plot = "pairs")
#featurePlot(x=mtcars[,-1], y = mtcars$mpg, plot = "pairs")
pairs(mpg ~ ., data= mtcars)
```

###MPG vs. transmission type
```{r appendix-fig2, fig.cap="Figure 2. MPG VS. Transmission Type", echo=FALSE}
#plot(mtcars$mpg ~ ., data = mtcars)
#plot(factor(mtcars$am), mtcars$mpg,data=mtcars)
pp <- ggplot(mtcars, aes(am, mpg))
pp + geom_boxplot(aes(fill= am)) + geom_jitter()
```

###Residuals plot
```{r appendix-fig3, fig.cap="Figure 3. Residuals", echo=FALSE}
plot(bestModel)
```   

###Exploration of other linear models
```{r echo=FALSE}
qq <- qplot(hp, mpg, colour = factor(am), data=mtcars)
qq + geom_smooth(method='lm', formula = y~x)

# qq <- qplot(cyl, mpg, colour = factor(am), data=mtcars)
# qq + geom_smooth(method='lm', formula = y~x, aes(group = 2)) 

qq <- qplot(wt, mpg, colour = factor(am), data=mtcars)
qq + geom_smooth(method='lm', formula = y~x)

# qq <- qplot(qsec, mpg, colour = factor(am), data=mtcars)
# qq + geom_smooth(method='lm', formula = y~x)

# qq <- qplot(vs, mpg, colour = factor(am), data=mtcars)
# qq + geom_smooth(method='lm', formula = y~x, aes(group = 1))

qq <- qplot(carb, mpg, colour = factor(am), data=mtcars)
qq + geom_smooth(method='lm', formula = y~x, aes(group = 1))

```   